Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-12 infercnv/reports/gains_and_losses.Rmd
reorganise gains_and_losses report

First, we load the CITE-seq object. We convert it to a SingleCellExperiment object.

@@

  # generate coldata
  colData <-
    obj$meta_data[cell_order, ]

  # populate sce
  sce <- 
    SingleCellExperiment::SingleCellExperiment(
      assays = assays,
      colData = colData
    )

@@

@@

Next, we add the CNV metadata from the inferCNV run to the object.

```{r get_infercnv}
sce <- xfun::cache_rds({

  # get cnv metadata, reorder, left join to colData
  cnvs <- 
    list.files(
      infercnv_dir,
      pattern = 'map_metadata_from_infercnv.txt',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(infercnv_dir, file) %>% 
        read.table(sep = '\t')
      }) %>%
    dplyr::bind_rows()

  # left join cnvs to coldata, reorder, restore rownames
  colData <- merge(colData, cnvs, by = 'row.names', all.x = T)
  rownames(colData) <- colData[, 1]
  colData <- colData[, -1]
  colData <- colData[colnames(sce), ]

  # add to sce
  sce@colData <- colData

}, file = 'sce_w_infercnv.rds')
```

@@

# split gain and loss
p_dat <-
  c('dupli', 'loss') %>% 
  purrr::set_names() %>%
  purrr::map(function(direction) {
    list(
      matrix = 
        p_dat %>% 
        dplyr::select(dplyr::contains(paste0('proportion_', direction))) %>%
        dplyr::filter(!is.na(paste0('proportion_', direction, '_chr1'))) %>%
        as.matrix(),
      annot = p_dat %>% dplyr::select(sample, cell_type)
    )
  })

# plot heatmaps
pheatmap::pheatmap(p_dat$dupli$matrix, cluster_cols = F, cluster_rows = F)
pheatmap::pheatmap(sce@colData[,grepl('proportion_dupli', colnames(sce@colData))])

# function: plot gains and losses
plot_gains_and_losses <- 
  function(cnvs, annot) {
    cnvs %>%
      dplyr::rename(annotation = tidyselect::all_of(annot)) %>%
      dplyr::filter(measure == 'proportion', event != 'cnv') %>%
      dplyr::group_by(id, chr, event, annotation) %>%
      dplyr::summarise(proportion = mean(value)) %>%
      ggplot(aes(x = chr, y = id, alpha = proportion, fill = event)) +
      geom_tile() +
      theme_classic() +
      theme(panel.spacing = unit(0, "lines"),
            axis.text.y = element_text(size = 3),
            axis.text.x = element_text(angle = 270, hjust = 0, vjust = 0.5)) +
      scale_fill_manual(values = c('loss' = scales::muted('blue'),
                                   'dupli' = scales::muted('red'))) +
      scale_alpha(range = c(0, 1)) +
      facet_grid(annotation ~ event)
  }

plot_gains_and_losses(cnvs, 'cell_type')
plot_gains_and_losses(cnvs, 'cluster_name')
```

```{r plot_umaps, fig.height = 15}
list(
  # plot celltype
  cnvs %>%
    dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
    ggplot(aes(x = UMAP1, y = UMAP2, colour = cell_type)) +
    geom_point(size = 0.05) +
    theme_classic() +
    coord_fixed(),
  # plot chr7 gain on umap
  cnvs %>%
    dplyr::filter(measure == 'proportion', event == 'dupli', chr == 'chr7') %>%
    dplyr::arrange(value) %>%
    ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) +
    geom_point(size = 0.05) +
    theme_classic() +
    scale_colour_viridis_c() +
    coord_fixed() +
    ggtitle('proportion of genes duplicated on chr7')
) %>%
  patchwork::wrap_plots(ncol = 1)
```

@@

We will define cells with chr7 gain as those affecting >90% of genes.

```{r pseudobulk_chr7_gain}
pseudobulk <- xfun::cache_rds({

@@

  # pseudobulk the sce object by chr7 status
  reduced_sce <-
    glmGamPoi::pseudobulk_sce(
      sce, 
      group_by = vars(chr7_gain, cell_type)
    )

}, file = 'pseudobulk.rds')
```

>>>>

First, we load the CITE-seq object. We convert it to a SingleCellExperiment object. Next, we add the CNV metadata from the inferCNV run to the object.

@@

  # generate coldata
  col_data <-
    obj$meta_data[cell_order, ]

  # populate sce
  sce <- 
    SingleCellExperiment::SingleCellExperiment(
      assays = assays,
      colData = col_data
    )

@@

  # get cnv metadata
  cnvs <- 
    list.files(
      infercnv_dir,
      pattern = 'map_metadata_from_infercnv.txt',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(infercnv_dir, file) %>% 
        read.table(sep = '\t')
      }) %>%
    dplyr::bind_rows()

  # left join cnvs to coldata, reorder, restore rownames
  col_data <- merge(col_data, cnvs, by = 'row.names', all.x = T)
  rownames(col_data) <- col_data[, 1]
  col_data <- 
    col_data[colnames(sce), -1] %>%
    DataFrame()

  # add to sce
  sce@colData <- col_data

@@

@@

# split gain and loss
p_dat <-
  c('dupli', 'loss') %>% 
  purrr::set_names() %>%
  purrr::map(function(direction) {
    p_dat_i <-
      p_dat %>% 
        dplyr::filter(!is.na(paste0('proportion_', direction, '_chr1')))
    list(
      matrix = 
        p_dat_i %>%
        dplyr::select(dplyr::contains(paste0('proportion_', direction))) %>%
        as.matrix(),
      annot = 
        p_dat_i %>% 
        dplyr::select(sample, cell_type)
    )
  })

# plot heatmaps
pheatmap::pheatmap(p_dat$dupli$matrix, cluster_cols = F, cluster_rows = F)
pheatmap::pheatmap(sce@colData[,grepl('proportion_dupli', colnames(sce@colData))])
```

@@

We will define cells with chr7 gain as those affecting >90% of genes. We will pseudobulk by patient, celltype, and chr7 ploidy to produce a reduced object.

```{r pseudobulk_chr7_gain}
reduced_sce <- xfun::cache_rds({

@@

  # pseudobulk the sce object by patient, celltype, and chr7 ploidy
  reduced_sce <-
    glmGamPoi::pseudobulk_sce(
      sce, 
      group_by = vars(sample, cell_type, chr7_gain),
      n_cells = n()
    )

  # remove NAs
  reduced_sce <- reduced_sce[, !is.na(reduced_sce$cell_type)]

  # return
  reduced_sce

}, file = 'reduced_sce.rds')
```

With the reduced data, we can conduct differential expression analysis, treating the data like bulk RNA-seq data.

```{r deseq, message = T}
# use DESeq2's size factor calculation procedure
fit <-
  glmGamPoi::glm_gp(
    reduced_sce,
    design = ~ chr7_gain * cell_type + sample,
    size_factor = 'ratio',
    verbose = T
  )
res <-
  glmGamPoi::test_de(
    fit, 
    contrast = 
      cond(cell_type = 'Stromal cell', chr7_gain = 1) -
      cond(cell_type = 'Stromal cell', chr7_gain = 0)
  )
```

```{r plot_de}
res %>%
  ggplot(aes(x = lfc, y = log10(pval))) +
  geom_point(aes(colour = adj_pval < 0.01), size = 0.5) +
  t
```
